---
title: "Lab 3: COVID-19"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: https://github.com/SarahRoseCulhane
    email: sculhane@colostate.edu
format: html
 theme: Solar
---

**question 1: Public Data**

```{r}
library(tidyverse); library(flextable)
data = read_csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv')
```

Here is how I created an object called my.date set as "2022-02-01" and an object called my.state, set as "Colorado"

```{r}
my.date <- as.Date("2022-02-01")  
my.state <- "Colorado"
```

**Question 2: Daily Summary**

Here, I created a subset limiting the data to Colorado and adding new columns for new cases and new deaths

```{r}
colorado_covid <- data %>%
  filter(state == my.state) %>%
  arrange(county, date) %>%
  group_by(county) %>%
  mutate(
    new_cases = cases - lag(cases, default = 0),  
    new_deaths = deaths - lag(deaths, default = 0)  
  ) %>%
  ungroup()

```

Here is the code I used to generate a data set of the 5 counties with the most cumulative cases.

```{r}
covid_most_cumulative <- colorado_data %>%
  filter(date == my.date) %>%
  arrange(desc(cases)) %>%
  slice_head(n = 5) %>%
  select(county, cases) 
```

Here is the code I used to make a table of the data using the flextable function.

```{r}
covid_most_cumulative_table <- flextable(covid_most_cumulative) %>%
  add_footer_lines("Top 5 counties in Colorado with the most cumulative COVID cases on 2022-02-01") %>%
  set_caption("Table 1: Most cumulative COVID cases") %>%
  set_table_properties(layout = "autofit", width = 0.5) %>%
  align(align = "center", part = "all") 

covid_most_cumulative_table
```

Here is the code I used to generate a table showing the top 5 counties with the most new cases

```{r}
covid_new_cases <- colorado_data %>%
  filter(date == my.date) %>%
  select (county, new_cases) %>%
  arrange(desc(new_cases)) %>%
  slice_head(n = 5) 
```

Here is the code I used to generate a table of the data using flextable

```{r}
covid_new_cases_table <- flextable(covid_new_cases) %>%
  add_footer_lines("Top 5 counties in Colorado with the most new cases on 2022-02-01") %>%
  set_caption("Table 2: Most new COVID cases") %>%
  set_table_properties(layout = "autofit", width = 0.5) %>%
  align(align = "center", part = "all")
covid_new_cases_table

```

**Question 3: Normalizing Data** I used the code to complete the following tasks:

I used the following code to read the population data
```{r}
pop_url <- 'https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv'
```

First, I renamed the "fips" column "FIP" in the covid data
```{r}
colorado_covid_temp <- colorado_covid %>%
  rename(FIP = fips)
colorado_covid_temp
```
Then, I loaded the population data
```{r}
population_data = read_csv('https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv')
```

I used the following code to:
-   Converted the STATE numeric into a character forced to 2 digits with a leading 0 (when needed)
-   Converted the COUNTY numeric into a character forced to 3 digits with leading 0’s (when needed)
-   Created a FIP variable the STATE numeric into a character forced to 2 digits with a leading 0 (when needed)
```{r}
#Converted STATE numeric and COUNTY numberic into characters
population_data$STATE_FIP <- sprintf("%02s", population_data$STATE)
population_data$COUNTY_FIP <- sprintf("%03s", population_data$COUNTY)

#created a FIP variable
population_data$FIP <- paste0(population_data$STATE_FIP, population_data$COUNTY_FIP)
```

Here is the code I used to select only columns that contain "NAME" or "2021" and remove counties with FIP code "000"

```{r}
population_data_2021 <- population_data_filtered %>%
  filter(COUNTY_FIP !="000", STNAME == my.state) %>%
   select(contains("NAME") | contains("2021"), FIP)
```

Here is the code I used to find the range of populations in Colorado counties in 2021. The range was between 741 and 737287 people
```{r}
range(population_data_2021$POPESTIMATE2021)
```
Here, I filtered the covid data to only contain the data for 2021.
```{r}
colorado_covid_2021 <- colorado_covid_temp %>%
  filter(year(date) == 2021) %>%     
  select(date, everything())  
  
```
Here, joined colorado_covid_2021 and poulation_data_2021
```{r}
per_capita_2021 <- colorado_covid_2021 %>%
  left_join(population_data_2021, by = "FIP")
```

Here, I calculated per capita cumulative cases, per capita new cases, and per capita new deaths

```{r}
per_capita_2021 <- per_capita_2021 %>%
  mutate(
    per_capita_cumulative_cases = cases / POPESTIMATE2021,  
    per_capita_new_cases = new_cases / POPESTIMATE2021,               per_capita_new_deaths = new_deaths / POPESTIMATE2021)    
```
However, it looks like there are a lot of duplicates, so here is the code I used to clean up the dataset.
```{r}
per_capita_2021_clean <- per_capita_2021 %>%
  distinct()  
```

Here is the code I used calculate the total per capita new cases and total per capita new cases, and per capita cumulative cases per county

```{r}
per_capita_county_summary <- per_capita_2021_clean %>%
  group_by(county) %>%  
  summarize(
    total_cumulative_cases = sum(per_capita_cumulative_cases, na.rm = TRUE),  
    total_new_cases = sum(per_capita_new_cases, na.rm = TRUE),    
    total_new_deaths = sum(per_capita_new_deaths, na.rm = TRUE))
```

Here is the code I used to generate a table with the most cumulative cases per capita

```{r}
per_capita_county_summary_table <- per_capita_county_summary %>%
  select (county, total_cumulative_cases) %>%
  arrange(desc(total_cumulative_cases)) %>%
  slice_head(n = 5) 

per_capita_county_summary_table <- flextable(per_capita_county_summary_table) %>%
  add_footer_lines("Top 5 counties in Colorado with the most cumulative cases per capita in 2021") %>%
  set_caption("Table 3: cumulative cases per capita") %>%
  set_table_properties(layout = "autofit", width = 0.5) %>%
  align(align = "center", part = "all")
per_capita_county_summary_table
```
Here is the code I used to create a table of the top five counties in colorado with the most new cases in 2021
```{r}
per_capita_county_new_cases <- per_capita_county_summary %>%
  filter(date == my.date) %>%
  select (county, total_new_cases) %>%
  arrange(desc(total_new_cases)) %>%
  slice_head(n = 5) 

per_capita_county_new_cases <-  flextable(per_capita_county_new_cases) %>%
  add_footer_lines("Top 5 counties in Colorado with the most new cases per capita in 2021") %>%
  set_caption("Table 3: new cases per capita") %>%
  set_table_properties(layout = "autofit", width = 0.5) %>%
  align(align = "center", part = "all")
per_capita_county_new_cases
```
Question 4: Rolling Thresholds

Here is how I filtered the COVID/population data to only include the last 14 days
```{r}
covid_last_14_days <- per_capita_2021_clean %>%
  
```






